Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-07-24 submission/manuscript.Rmd
discussion draft

We found this dysregulation in tissue identity likely begins during early emrbyogenesis, as germline genes and other tissue-enriched genes are aberrantly expressed in _Kdm5c_-KO epiblast-like cells (EpiLCs), an _in vitro_ model of the post-implantation embryo. _In vivo_, germline genes are typically decommissioned in epiblast stem cells and remain silenced as the epiblast differentiates into the body's somatic tissues[]. However, a small subset of epiblast stem cells will receive signals to reactivate germline gene expression to become the primoridal germ cells (PGCs) that will ultimately form the mature germline[]. This process can be mimicked _in vitro_ by differentiating EpiLCs into primordial germ cell-like cells (PGCLCs)[XXX]. Thus, misexpression of germline genes in _Kdm5c_-KO EpiLCs might suggest they are progressing beyond EpiLC differentiation and becoming PGCLCs. Yet, _Kdm5c_-KO EpiLCs had proper expression of EpiLC marker genes for primed pluripotency and we observed no difference in cellular morphology during ESC to EpiLC differentiation. Furthermore, we saw no significant change in _Otx2_ expression, an epiblast stem cell marker that is known to repress epiblast and EpiLC differentiation into PGC and PGCLCs[@zhangOTX2RestrictsEntry2018]. This, in conjunction with _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to germline identity.

- Only a small number of germline genes were shared between _Kdm5c_-KO EpiLCs and in the brain.

dysregulation of tissue identity

- In line with previous work on other chromatin regulators, we found dysregulation of _Kdm5c_-KO tissue identity began during early embryogenesis. While _Kdm5c_-KO epiblast-like cells (EpiLCs)

- KDM5C represses germline genes during early embryogenesis
	- Found this dysregulation in tissue identity begins during _Kdm5c_-KO early embryogenesis
	- EpiLC express early regualtors, Brain late sperm - program seems to be continuing across development
	- Importantly, even though germline genes are misexpressed, markers of EpiLC differentiation and EpiLC morphology is unimpaired. 
		- Epiblast can become primoridal germ cells
		- Otx2 is expressed in EpiLCs and is known to repress PGC identity. It's properly expressed in _Kdm5c_-KO EpiLCs, further supporting they aren't just becoming PGCs https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6485399/ 
		- Germline Program happening in the background of typical development.

- One of the genes misexpressed in _Kdm5c_-KO EpiLCs is Dazl, _Deleted in azoospermia like (Dazl)_, a key regulator of germline development that promotes the translation of germline-specific transcripts[XXX].
	- 2 other studies of dazl regulators in ESCs also found KDM5C represses dazl. 
	- While crucial for germ cell development, Dazl is also expressed in ESCs and at the 2-cell stage. 
	- Some genes important for germ cell, ESC, and 2-cell development given they are important for self-renewal.  
	- We did not observe dysregulation of 2-cell-specific regulators like Dux. 
	- We found Dazl was expressed in _Kdm5c_-KO EpiLCs, when _Dazl_ it typically decommissioned. This indicates KDM5C loss results in long-term ectopic expression.
	- We additionally found DAZL protein is ectopically expressed and localized to the cytoplasm. Given that a significant number of KDM5C-KO gemrline DEGs are DAZL targets, it may promote the ectopic germline program.

	- Very recently two other studies identified KDM5C represses Dazl in ESCs. 
	- - KDM5C was very recently shown to repress _Deleted in azoospermia like (Dazl)_, a key regulator of germline development, in mouse embryonic stem cells (ESCs)[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023]. However, KDM5C binding and _Kdm5c_-KO germline gene misexpression has yet to be globally characterized during early embryogenesis. Given that _Dazl_ and other germline-enriched genes can also be expressed in ESCs and at the 2-cell stage, it is unclear if KDM5C has a direct role in the long-term germline gene silencing that occurs in the post-implantation epiblast.

- We globally characterized KDM5C's repression of germline genes during early embryogenesis and in mature neurons. 
	- In line with previous work, we found KDM5C did not regulate germline genes in neurons. 
	- KDM5C-bound genes in EpiLCs were enriched for germline ontologies, suggesting a major role of KDM5c during embryogenesis
	- While KDM5C directly binds some germline genes, including _Dazl_, many of the genes dysregulated in _Kdm5c_-KO were not direct targets. 
		- Stra8 notable exception
		- Brain genes, including late-stage spermatogenesis genes
		- Suggests germline genes can be dysregulated direct and indirect of KDM5C resgulation
		- Further supports germline programs can be ectopically activated
		- Talk about different motifs if we do see differences there and if it explains direct vs indirect dysregulation
			- Could also mention PRC1.6/KDM5C binds pcgf6 in ESCs

@@

@@

- While a significant portion of KDM5C-bound promoters are germline genes, many germline genes expressed during _Kdm5c_-KO embryogenesis are not directly bound by kdm5c. 
- One notable example is Stra8, a transcription factor important for germ cell specific transcription and meiotic initiation
-

While Directly binds to a subset of germline genes during early embryogenesis 
- KDM5C binds to a subset of germline genes during early embryogenesis

@@

- Altogether, these data indicate that while some germline genes are misexpressed due to direct loss of KDM5C binding during emryogenesis, secondary downstream mechanisms can also promote their aberrant transcription.  
	- This shift from meiotic genes to later spermatogenesis genes in the hippocampus suggests the germline developmental program could occur ectopically as _Kdm5c_-KO cells progresses through somatic tissue development. **note: this is strengthened by the ChIP-seq data since KDM5C is not directly bound to many brain/flagellar DEGs.This point might be stronger in the ChIPseq figure**

>>>>

Dysregulation of _Kdm5c_-KO tissue identity begins during early emrbyogenesis, as germline and other tissue-enriched genes are aberrantly expressed in epiblast-like cells (EpiLCs). _In vivo_, germline genes are typically decommissioned in epiblast stem cells and remain silenced as the epiblast differentiates into the body's somatic tissues[]. However, a small subset of epiblast stem cells will receive signals to reactivate germline gene expression to become the primoridal germ cells (PGCs) that will ultimately form the mature germline[]. This process can be mimicked _in vitro_ by differentiating EpiLCs into primordial germ cell-like cells (PGCLCs)[XXX]. Thus, misexpression of germline genes in _Kdm5c_-KO EpiLCs might suggest they are progressing beyond EpiLC differentiation and becoming PGCLCs. Yet, _Kdm5c_-KO EpiLCs had proper expression of EpiLC marker genes for primed pluripotency and we observed no difference in cellular morphology during ESC to EpiLC differentiation. Furthermore, we saw no significant change in _Otx2_ expression, an epiblast stem cell marker that is known to repress epiblast and EpiLC differentiation into PGC and PGCLCs[@zhangOTX2RestrictsEntry2018]. This, in conjunction with _Kdm5c_-KO mice being viable, suggests germline gene expression is occuring ectopically in conjunction with typical developmental programs, rather than a complete shift to germline identity.

- change first sentence: Other chromatin regulators involved in germline gene repression act in ESCs and EpiLCs. We observed this is true for KDM5C
We then globally characterized KDM5C binding at germline-enriched gene promoters through analysis of KDM5C ChIP-seq in EpiLCs and primary hippocampus and cortex neuron cultures (PNCs). While we observed high enrichment of KDM5C peaks at germline genes in EpiLCs, KDM5C was not bound to germline gene promoters in PNCs. This suggests dysregulation of germline genes in the mature _Kdm5c_-KO brain is due to loss of repression during embryogenesis, which is consistent with previous work that found introducing human KDM5C into _Kdm5c_-KO PNCs does not repress germline transcripts[@scandagliaLossKdm5cCauses2017]. Although KDM5C is enriched at germline gene promoters in EpiLCs, KDM5C was only bound to about a third of EpiLC germline-enriched DEGs. One notable DEG not bound by KDM5C is _Stra8_, a transcription factor activated by retinoic acid signaling in germ cells that promotes meiotic initiation[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006]. Retinoic acid can only activate _Stra8_ expression when DAZL is present, which is a direct target of KDM5C and is aberrantly transcribed and translated in _Kdm5c_-KO EpiLCs. This indicates germline genes can be aberrantly expressed in _Kdm5c_-KO cells indirectly of KDM5C regulation through activation by other ectopic germline programs. Consistent with this idea, many _Kdm5c_-KO EpiLC germline DEGs are important for early germ cell development and meiosis while those expressed in the mature _Kdm5c_-KO brain are involved in late sperm development. Altogether, this indicates ectopic germline programs are, to some extent, progressing through germ cell developmental stages over the course of _Kdm5c_-KO development.

- Talk about different motifs if we do see differences there and if it explains direct vs indirect dysregulation
	- Could also mention PRC1.6/KDM5C binds pcgf6 in ESCs
- Maybe talk about other regulators/if anything is known about them long term? Or save for 2nd paper.
	- another germline repressor: https://pubmed.ncbi.nlm.nih.gov/35345626/
- Might be good to look at for retinoic acid paper (WT germ expression): https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0005654

It is important to note that while we focused on KDM5C's regulation of germline genes, some germline-enriched genes are also expressed in na&#239;ve ESCs and at the 2-cell stage due to their roles in pluripotency and self-renewal. For example, although primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific RNAs, _Dazl_ is also expressed in na&#239;ve ESCs _in vitro_[@wellingDAZLRegulatesTet12015], the inner cell mass _in vivo_[@wellingDAZLRegulatesTet12015], and at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], but is silenced when cells transition from na&#239ve to primed pluripotency during ESC to EpiLC differentiation[@mochizukiRepressionGermlineGenes2021]. Very recently, two screens of _Dazl_-repressors in ESCs identified KDM5C as a direct regulator of _Dazl_ expression[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023]. Interestingly, one screen found _Kdm5c_-KO ESCs also aberrantly expressed 2-cell-specific genes, indicating KDM5C negatively regulates totipotency in ESCs[@guptaGenomewideScreenReveals2023]. We found that while KDM5C also repressess _Dazl_ expression in EpiLCs, _Kdm5c_-KO EpiLCs do not express 2-cell specific genes like _Dux_ and _Zscan4c_. Out of the four 2-cell regulators characterized in ESCs, KDM5C was the only factor whose repression of _Dazl_ was independent of _Dux_ expression[@guptaGenomewideScreenReveals2023]. Together, shis suggests the 2-cell-like state in _Kdm5c_-KO ESCs is an indirect consequence of KDM5C's primary role in germline gene repression, including germline genes that are also involved in pluripotency and self-renewal.

@@

- We additionally found _Kdm5c_-KO EpiLCs ectopically express DAZL protein that is localized to the cytoplasm, similar to its morphology in spermatogonia[@mikedisDAZLMediatesBroad2020].

- end with something like: this indicates kdm5c not only modulates pluripotency and self-renewal in ESCs, but also has a role in long-term silencing of germline genes 
	- then transition into the long term silencing mechanism paragraph

@@

- While a significant portion of KDM5C-bound promoters are germline genes, many germline genes expressed during _Kdm5c_-KO embryogenesis are not directly bound by kdm5c. 
- One notable example is Stra8, a transcription factor important for germ cell specific transcription and meiotic initiation

@@

- Altogether, these data indicate that while some germline genes are misexpressed due to direct loss of KDM5C binding during emryogenesis, secondary downstream mechanisms can also promote their aberrant transcription.
